Dependently Typed Languages in Statix

### 2023-01-16 paper-evcs/paper.tex
Changes based on Aron's feedback v2

@@

Spoofax is a language workbench: \label{key}a collection of tools that enable the development of textual languages~\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. It is a declarative language that uses inference rules to define the static semantics. From these rules a type checker and other editor tools can be derived. Statix aims to cover a broad range of languages and type systems. However, no attempts have been made to express dependently typed languages in Statix until now.

@@

The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}. One extra feature was added that is not present in the Calculus of Constructions: let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The concrete syntax of the language is available in figure \ref{fig:syntax}.

@@

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (_: Type -> Type) (\x: Type. x)
\end{lstlisting}

@@

To type check the base language, we need to store information about the names that are in scope at each point in the program. There are two different cases, names that do not have a known value (only a type), which are names created by function arguments, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}

@@

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} { 
			\bhr
			{ \scope{s}{\co{Type}()} }
			{ [] }
			{ \scope{s}{\co{Type}()}} 
		}

@@

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} {
			\toe{s}{\co{Type}()}{\co{Type}()}
		}

		% Let rule
		\inferrule{ 
			\toe{s}{v}{t_e}
			\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
		}{
			\toe{s}{\co{Let}(x, e, b)}{t_b}
		}

@@

		% FnType rule
		\inferrule{ 
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}()} 
			\\ \bred{s}{a}{a'}
			\\\\  \toe{\co{sPutType}(s, x, a')}{b}{t_b}
			\\ \beq{t_b}{\co{Type}()}
		} { 
			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}()}
		}

@@

	\end{mathpar}
	\caption{Rules for type checking the Calculus of Constructions}
	\label{fig:type-check-rules}
\end{figure}

@@

\paragraph*{Inductive Data Types}
Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types. We can also generate eliminators for these data types to use their values. All of this has been implemented in Statix. 
\begin{lstlisting}
data Maybe (T : Type) = 
    None : Maybe T,
    Some : _: T -> Maybe T;
\end{lstlisting}

>>>>

\makeatletter
\newcommand\@eatpar{\@ifnextchar\par{\expandafter\@eatpar\@gobble}\relax}
\newcommand{\figuresection}[2][]{%
	\par%
	{\sffamily\bfseries #2}\hfill{#1}%
	\smallskip%
	\@eatpar}
\makeatother

@@

Spoofax is a language workbench: \label{key}a collection of tools that enable the development of textual languages~\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. It is a declarative language that uses inference rules to define static semantics. From these rules a type checker and other editor tools can be derived. Statix aims to cover a broad range of languages and type systems. However, no attempts have been made to express dependently typed languages in Statix until now.

@@

The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}. One extra feature was added that is not present in the Calculus of Constructions: let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The concrete syntax~\cite{sdf3} of the language is available in figure \ref{fig:syntax}.

Please observe that the syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language. Furthermore, \verb|FnType| assigns a name to its first argument to allow the return type of the function to depend on the argument type.

@@

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (T: Type -> Type) (\x: Type. x)
\end{lstlisting}

@@

To type check the base language, we need to store information about the names that are in scope at each point in the program. There are two different cases, names that do not have a known value (only a type), which are names created by function arguments and dependent function types, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}

@@

\begin{figure}[h]
	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}

	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} { 
			\bhr
			{ \scope{s}{\co{Type}()} }
			{ [] }
			{ \scope{s}{\co{Type}()}} 
		}

@@

\begin{figure}[h]
	\figuresection[\fbox{$\toe{s}{e}{t}$}]{Type checking rules}
	\begin{mathpar}	
		\mprset{vskip=0.7ex}
		% Type rule
		\inferrule{ 
		} {
			\toe{s}{\co{Type}()}{\co{Type}()}
		}

		% Let rule
		\inferrule{ 
			\toe{s}{e}{t_e}
			\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
		}{
			\toe{s}{\co{Let}(x, e, b)}{t_b}
		}

@@

		% FnType rule

		\inferrule{ 
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}()} 
			\\ \bred{s}{a}{a'}
			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
			\\ \beq{t_b}{\co{Type}()}
		} { 
			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}()}
		}

@@

		\mprset{vskip=0ex}
	\end{mathpar}
	\caption{Rules for type checking the Calculus of Constructions}
	\label{fig:type-check-rules}
\end{figure}

@@

\paragraph*{Inductive Data Types}
Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types. We can also generate eliminators for these data types to use their values. All of this has been implemented in Statix. 
\begin{lstlisting}
data Maybe (T : Type) = 
    None : Maybe T,
    Some : x: T -> Maybe T;
\end{lstlisting}
